---
title: "Multiple linear regression"
subtitle: "Types of predictors"
date: "01-27-2026"
date-format: "MMMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 06-predictor-type-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
filters: 
  - webr
webr: 
  packages: ['tidyverse', 'broom', 'openintro']
---

## Announcements

-   Lab 02 due **TODAY** at 11:59pm

-   HW 01 due **Thursday, January 29** at 11:59pm

-   (Optional) Submit team request by T**hursday, January 29 at 5pm**: <https://forms.office.com/r/3WHjriZjM9>

## Topics

-   Categorical predictors

-   Centering quantitative predictors

-   Standardizing quantitative predictors

-   Interaction terms

## Computing setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(viridis) #adjust color palette

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

## Data: Peer-to-peer lender

Today's data is a sample of 50 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

```{r}
#| echo: false

loan50 <- loan50 |>
  mutate(annual_income_th = annual_income / 1000)

loan50 |>
  select(annual_income_th, debt_to_income, verified_income, interest_rate)

# get rid of the hidden blank column of verified income 
loan50$verified_income <- droplevels(loan50$verified_income)
```

## Variables

**Predictors**:

::: nonincremental
-   `annual_income_th`: Annual income (in \$1000s)
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
:::

**Response**: `interest_rate`: Interest rate for the loan

## Response vs. predictors {.small}

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 10
p4 <- ggplot(loan50, aes(x = verified_income, y = interest_rate)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    y = "Interest rate",
    x = "Income verification"
  )

p5 <- ggplot(loan50, aes(x = debt_to_income, y = interest_rate)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Debt to income ratio"
  )


p6 <- ggplot(loan50, aes(x = annual_income_th, y = interest_rate)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Annual income (in $1000s)"
  )

p4 + p5 / p6
```

**Goal**: Use these predictors in a single model to understand variability in interest rate.

## Model fit in R

```{r}
#| echo: true

int_fit <- lm(interest_rate ~ debt_to_income + verified_income  + annual_income_th,
              data = loan50)

tidy(int_fit) |>
  kable(digits = 3)
```

# Categorical predictors

## Matrix form of multiple linear regression

$$
\underbrace{
\begin{bmatrix}
y_1 \\
\vdots \\
y_n
\end{bmatrix} }_
{\mathbf{y}} \hspace{3mm}
= 
\hspace{3mm}
\underbrace{
\begin{bmatrix}
1 &x_{11} & \dots & x_{1p}\\
\vdots & \vdots &\ddots & \vdots \\
1 &  x_{n1} & \dots &x_{np}
\end{bmatrix}
}_{\mathbf{X}}
\hspace{2mm}
\underbrace{
\begin{bmatrix}
\beta_0 \\
\beta_1 \\
\vdots \\
\beta_p
\end{bmatrix}
}_{\boldsymbol{\beta}}
\hspace{3mm}
+
\hspace{3mm}
\underbrace{
\begin{bmatrix}
\epsilon_1 \\
\vdots\\
\epsilon_n
\end{bmatrix}
}_\boldsymbol{\epsilon}
$$

<center>How might we include a categorical predictor with $k$ levels in the design matrix, $\mathbf{X}$ ?</center>

## Indicator variables

-   Suppose there is a categorical variable with $k$ levels

-   We can make $k$ indicator variables from the data - one indicator for each level

-   An **indicator (dummy) variable** takes values 1 or 0

    -   1 if the observation belongs to that level

    -   0 if the observation does not belong to that level

## Indicator variables {.midi}

Suppose we want to predict the amount of sleep a Duke student gets based on whether they are in Pratt (Pratt Yes/ No are the only two options). Consider the model

$$
Sleep_i = \beta_0 + \beta_1\mathbf{1}(Pratt_i = \texttt{Yes}) + \beta_2\mathbf{1}(Pratt_i = \texttt{No})
$$

::: question
-   Write out the design matrix for this hypothesized linear model.

-   Demonstrate that the design matrix is not of full column rank (that is, affirmatively provide one of the columns in terms of the others).

-   Use this intuition to explain why when we include categorical predictors, we cannot include both indicators for every level of the variable *and* an intercept.
:::

## Indicator variables for `verified_income`

```{r}
#| echo: true

loan50 <- loan50 |>
  mutate(
    not_verified = factor(if_else(verified_income == "Not Verified", 1, 0)),
    source_verified = factor(if_else(verified_income == "Source Verified", 1, 0)),
    verified = factor(if_else(verified_income == "Verified", 1, 0))
  )
```

. . .

```{r}
#| echo: false
loan50 |>
  select(verified_income, not_verified, source_verified, verified) |>
  slice(1, 3, 6)
```

## Trying to use all indicators in R {.midi}

::: question
Run the code below to fit a model using `debt_to_income`, `annual_income_th`, and all the indicator variables for `verified_income` to predict `interest_rate`. What do you notice about the model output? Why did this happen?
:::

```{webr-r}
#| context: setup

loan50 <- loan50 |>
  mutate(
    not_verified = factor(if_else(verified_income == "Not Verified", 1, 0)),
    source_verified = factor(if_else(verified_income == "Source Verified", 1, 0)),
    verified = factor(if_else(verified_income == "Verified", 1, 0)),
    annual_income_th = annual_income / 1000) |>
  filter(verified_income %in% c("Not Verified", "Source Verified", "Verified"))

# get rid of the hidden blank column of verified income 
loan50$verified_income <- droplevels(loan50$verified_income)

```

```{webr-r}
interest_rate_fit <- lm(interest_rate ~ debt_to_income + annual_income_th + 
                        not_verified + source_verified + verified, 
                        data = loan50)

tidy(interest_rate_fit)
```

## Indicator variables in the model {.midi}

Given a categorical predictor with $k$ levels...

-   Use $k-1$ indicator variables in the model
-   The **baseline** is the category that doesn't have a term in the model
    -   This is also called the *reference level*
-   The coefficients of the indicator variables in the model are interpreted as the expected change in the response compared to the baseline, holding all other variables constant.

## AE: Model in R {.midi}

::: question
Now let’s take a look at the design matrix for the model with predictors `debt_to_income`, `annual_income_th`, and `verified_income`.

How does R choose the baseline level by default?
:::

```{webr-r}
# add code to obtain design matrix
```

## Exploring indicator variables {.small}

```{r}
#| echo: false
tidy(int_fit
, conf.int  = T) |>
  kable(digits = 3) |>
  row_spec(c(3,4), background = "#dce5b2")
```

::: question
What is the intercept for individuals with

-   Not verified income?

-   Source verified income?

-   Verified income?
:::

## Interpreting `verified_income` {.small}

```{r}
#| echo: false
tidy(int_fit
, conf.int  = T) |>
  kable(digits = 3) |>
  row_spec(c(3,4), background = "#dce5b2")
```

. . .

::: incremental
-   The baseline level is `Not verified`.
-   People with source verified income are expected to take a loan with an interest rate that is 2.211% higher, on average, than the rate on loans to those whose income is not verified, holding all else constant.
:::

## Centering

-   Centering a quantitative predictor means shifting every value by some constant $C$

-   One common type of centering is **mean-centering**, in which every value of a predictor is shifted by its mean

-   Only quantitative predictors are centered

-   Center all quantitative predictors in the model for ease of interpretation

::: question
What is one reason we might want to center the quantitative predictors? What are the units of centered variables?
:::

## Centering {.midi}

Use the `scale()` function with `center = TRUE` and `scale = FALSE` to mean-center variables

```{r}
#| output-location: fragment 

loan50 <- loan50 |>
  mutate(debt_to_inc_cent = scale(debt_to_income, center = TRUE, scale = FALSE), 
         annual_inc_cent = scale(annual_income_th, center = TRUE, scale = FALSE))

lm(interest_rate ~ debt_to_inc_cent + verified_income + annual_inc_cent, data = loan50) |> 
  tidy() |> kable(digits = 3)
```

## Centering

```{r}
#| echo: false
#| message: false
#| warning: false

orig_model <- tidy(int_fit) |>
  select(term, estimate) 

cent_model <- lm(interest_rate ~ debt_to_inc_cent + verified_income + annual_inc_cent, data = loan50) |> tidy() |>
select(estimate)

orig_model |> bind_cols(cent_model) |> 
  kable(digits = 3, 
        col.names = c("Term", "Original Model", "Centered Model"))
```

::: question
How has the model changed? How has the model remained the same?
:::

## Standardizing

-   Standardizing a quantitative predictor mean shifting every value by the mean and dividing by the standard deviation of that variable

-   Only quantitative predictors are standardized

-   Standardize all quantitative predictors in the model for ease of interpretation

::: question
What is one reason we might want to standardize the quantitative predictors? What are the units of standardized variables?
:::

## Standardizing {.midi}

Use the `scale()` function with `center = TRUE` and `scale = TRUE` to standardized variables

```{r}
#| output-location: fragment 
#| 
loan50 <- loan50 |>
  mutate(debt_to_inc_std = scale(debt_to_income, center = TRUE, scale = TRUE), 
         annual_inc_std = scale(annual_income_th, center = TRUE, scale = TRUE))

lm(interest_rate ~ debt_to_inc_std + verified_income + annual_inc_std, data = loan50) |>
  tidy() |> kable(digits = 3)
```

## Standardizing

```{r}
#| echo: false
#| warning: false
#| message: false

orig_model <- tidy(int_fit) |>
  select(term, estimate) 

std_model <- lm(interest_rate ~ debt_to_inc_std + verified_income + annual_inc_std, data = loan50) |> tidy() |>
select(estimate)

orig_model |> bind_cols(std_model) |> 
  kable(digits = 3, 
        col.names = c("Term", "Original Model", "Standardized Model"))
```

::: question
How has the model changed? How has the model remained the same?
:::

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Interest rate vs. annual income

The lines are not parallel indicating there is a potential **interaction effect**. The slope of annual income differs based on the income verification.

```{r}
#| echo: false
#| fig-align: center
#| warning: false
#| message: false

p1 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Annual income (in $1000s)",
    y = "Interest rate"
  )

p2 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate,
                 color = verified_income)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Annual income (in $1000s)", y = NULL, color = NULL) +
  theme(legend.position = c(0.6, 0.9)) +
  scale_color_viridis_d(end = 0.9)

p1 + p2 +
  plot_annotation(title = "Interest rate vs. annual income")
```

## AE: Model with interaction effect {.midi}

::: question
Fit the model with the predictors `debt_to_income`, `annual_income_th`, `verified_income` , and the interaction between `annual_income_th` and `verified_income`.

-   Write the estimated regression equation for the people with `Not Verified` income.

-   Write the estimated regression equation for people with `Verified` income.
:::

```{webr-r}
# code to fit and display model with interaction
```

## Interaction term in model {.smaller}

```{r}
#| echo: true
int_fit_2 <- lm(interest_rate ~ debt_to_income + verified_income + annual_income_th + verified_income * annual_income_th,
      data = loan50)
```

```{r}
#| echo: false
tidy(int_fit_2) |>
  kable(digits = 3) |>
  row_spec(c(6,7), background = "#dce5b2")
```

## Interpreting interaction terms

-   What the interaction means: The effect of annual income on the interest rate differs by -0.016 when the income is source verified compared to when it is not verified, holding all else constant.
-   Interpreting `annual_income` for source verified: If the income is source verified, we expect the interest rate to decrease by 0.023% (-0.007 + -0.016) for each additional thousand dollars in annual income, holding all else constant.

## Indicators and interactions

::: question
In general, how do

-   indicators for categorical predictors impact the model equation?

-   interaction terms impact the model equation?
:::

## Recap

-   Interpreted categorical predictors

-   Explored why the model includes $k-1$ terms for a categorical predictor with $k$ levels

-   Fit and interpreted models with centered and standardized variables

-   Interpreted interaction terms

## Next class

-   Inference for regression
-   Complete [Lecture 07 prepare](../prepare/prepare-lec07.html)
