---
title: "Multiple linear regression"
subtitle: "Types of predictors cont'd"
date: "02-03-2026"
date-format: "MMMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 06-predictor-types-contd-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements

-   Lab 03 due **TODAY**

    -   One submission per team

    -   Select every team member's name in Gradescope

-   [Statistics experience](../hw/stats-experience.html) due April 2

-   SSMU Mini DataFest - February 8

    -   See [Ed Discussion announcement](https://edstem.org/us/courses/90766/discussion/7577338)

## Topics

-   Interaction terms

## Computing setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(viridis) #adjust color palette

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

```{webr-r}
#| context: setup

loan50 <- loan50 |>
  mutate(
    not_verified = factor(if_else(verified_income == "Not Verified", 1, 0)),
    source_verified = factor(if_else(verified_income == "Source Verified", 1, 0)),
    verified = factor(if_else(verified_income == "Verified", 1, 0)),
    annual_income_th = annual_income / 1000) |>
  filter(verified_income %in% c("Not Verified", "Source Verified", "Verified"))

# get rid of the hidden blank column of verified income 
loan50$verified_income <- droplevels(loan50$verified_income)

```

## Data: Peer-to-peer lender

Today's data is a sample of 50 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

```{r}
#| echo: false

loan50 <- loan50 |>
  mutate(annual_income_th = annual_income / 1000)

loan50 |>
  select(annual_income_th, debt_to_income, verified_income, interest_rate)

# get rid of the hidden blank column of verified income 
loan50$verified_income <- droplevels(loan50$verified_income)
```

## Variables

**Predictors**:

::: nonincremental
-   `annual_income_th`: Annual income (in \$1000s)
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
:::

**Response**: `interest_rate`: Interest rate for the loan

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Interest rate vs. annual income

The lines are not parallel indicating there is a potential **interaction effect**. The slope of annual income differs based on the income verification.

```{r}
#| echo: false
#| fig-align: center
#| warning: false
#| message: false

p1 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Annual income (in $1000s)",
    y = "Interest rate"
  )

p2 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate,
                 color = verified_income)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Annual income (in $1000s)", y = NULL, color = NULL) +
  theme(legend.position = c(0.6, 0.9)) +
  scale_color_viridis_d(end = 0.9)

p1 + p2 +
  plot_annotation(title = "Interest rate vs. annual income")
```

## AE: Model with interaction effect {.midi}

::: question
:::

```{webr-r}
# code to fit and display model with interaction
```

## Model with interaction terms {.smaller}

```{r}
#| echo: true
int_fit_2 <- lm(interest_rate ~ debt_to_income + verified_income + annual_income_th + verified_income * annual_income_th,
      data = loan50)
```

```{r}
#| echo: false
tidy(int_fit_2) |>
  kable(digits = 3) |>
  row_spec(c(6,7), background = "#dce5b2")
```

## Model with interaction terms 

::: question
-   Write the estimated regression equation for the people with `Not Verified` income.

-   Write the estimated regression equation for people with `Verified` income.
:::

## Interpreting interaction terms

-   What the interaction means: The effect of annual income on the interest rate differs by -0.016 when the income is source verified compared to when it is not verified, holding all else constant.
-   Interpreting `annual_income` for source verified: If the income is source verified, we expect the interest rate to decrease by 0.023% (-0.007 + -0.016) for each additional thousand dollars in annual income, holding all else constant.

## Indicators and interactions

::: question
In general, how do

-   indicators for categorical predictors impact the model equation?

-   interaction terms impact the model equation?
:::

## Recap

-   Interpreted interaction terms
