---
title: "Multicollinearity"
author: "Prof. Maria Tackett"
date: "02-24-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 12-multicollinearity-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements

-   Project proposal due **TODAY** at 11:59pm on GitHub

-   February 25, 5-6 pm \@ East Duke 108: Duke Sports Analytics Club is hosting **Sam Marks**, Director of Business Strategy and Analytics at the Boston Bruins

    -   [Click here](https://docs.google.com/forms/d/e/1FAIpQLSdethCogomC5DTNwSyQy5N_R5fOkT5LfJGR27VSEHGfWIH-cw/viewform) to register

-   SSMU Mini DataFest #2: February 27 - March 1. [Click here](https://edstem.org/us/courses/90766/discussion/7720392) for announcement

-   March 20 - 22: ASA DataFest at Duke

    -   Learn more and register: <https://dukestatsci.github.io/datafest/#signup> 

## Computing set up

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)  
library(tidymodels)  
library(knitr)       
library(patchwork)
library(GGally)   # for pairwise plot matrix
library(corrplot) # for correlation matrix
# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Topics

-   Model conditions and diagnostics

-   Multicollinearity

    -   Definition

    -   How it impacts the model

    -   How to detect it

    -   What to do about it

# Model conditions and diagnostics

## Assumptions for regression {.midi}

$$
\mathbf{y}|\mathbf{X} \sim N(\mathbf{X}\boldsymbol{\beta}, \sigma_\epsilon^2\mathbf{I})
$$

1.  **Linearity:** There is a linear relationship between the response and predictor variables.
2.  **Equal variance:** The variability about the least squares line is generally constant.
3.  **Normality:** The distribution of the errors (residuals) is approximately normal.
4.  **Independence:** The errors (residuals) are independent from one another.

## Influential Point

An observation is **influential** if removing has a noticeable impact on the regression coefficients

```{r,echo=F}
set.seed(12)
n <- 20
x <- c(runif(n,0,1))
y <- 3*x + rnorm(n,0,.5)
new.pt <- data.frame(x=2,y=0)
x.new <- c(x,2)
y.new <- c(y,0)
data <- bind_cols(x=x.new,y=y.new)
p1<- ggplot(data=data,aes(x=x,y=y))+geom_point(alpha =0.5)  + 
              geom_point(data=new.pt,color="red",size=3) + 
  geom_smooth(method="lm",se=F) + 
  labs(title = "With Influential Point")+ theme_light()+
  theme(title=element_text(hjust=0.5,size=14)) + 
  scale_x_continuous(limits = c(0,2)) 

data2 <- bind_cols(x=x,y=y)
p2 <- ggplot(data=data2,aes(x=x,y=y))+geom_point(alpha=0.5) + geom_smooth(method="lm",se=F) + 
  labs(title="Without Influential Point") + 
  scale_x_continuous(limits = c(0, 2)) + theme_light() + theme(title=element_text(hjust=0.5,size=14))  
p1 + p2
```

## Cook's Distance

Cook's distance for the $i^{th}$ observation is

$$
D_i = \frac{r^2_i}{p + 1}\Big(\frac{h_{ii}}{1 - h_{ii}}\Big)
$$

where $r_i$ is the studentized residual and $h_{ii}$ is the leverage for the $i^{th}$ observation

This measure is a combination of

-   How well the model fits the $i^{th}$ observation

-   How far the $i^{th}$ observation is from the rest of the data

## Using Cook's Distance

-   An observation with large value of $D_i$ is said to have a strong influence on the predicted values

-   General thresholds .An observation with

    -   $D_i > 0.5$ is **moderately influential**

    -   $D_i > 1$ is **very influential**

## Lemurs data

```{r}
#| echo: false

library(viridis)

lemurs <- read_csv("data/lemurs-pcoq.csv") |>
  filter(age_at_wt_mo < 34)

ggplot(data = lemurs, aes(x = age_at_wt_mo, y = weight_g)) + 
  geom_point() + 
  labs(x = "Age in months",
       y = "Weight in grams", 
       title = "Weight vs. Age of PCOQ lemurs")

lemurs_fit <- lm(weight_g ~ age_at_wt_mo, data = lemurs)

lemurs_aug <- augment(lemurs_fit)
```

## Large leverage {.midi}

-   Measure of how far the predictor (combination of predictors) is from the typical value (combination)

-   The sum of the leverages for all points is $p + 1$, where $p$ is the number of predictors in the model . More specifically

    $$
    \sum_{i=1}^n h_{ii} = \text{rank}(\mathbf{H}) = \text{rank}(\mathbf{X}) = p+1
    $$

-   The average value of leverage, $h_{ii}$, is $\bar{h} =  \frac{(p+1)}{n}$

-   An observation has **large leverage** if $$h_{ii} > \frac{2(p+1)}{n}$$

## Scaled residuals

-   What is the best way to identify outlier points that don't fit the pattern from the regression line?

    -   Look for points that have large residuals

-   We can rescale residuals and put them on a common scale to more easily identify "large" residuals

-   We will consider two types of scaled residuals: standardized residuals and studentized residuals

## Standardized residuals

::: incremental
-   The variance of the residuals can be estimated by the mean squared residuals (MSR) $= \frac{SSR}{n - p - 1} = \hat{\sigma}^2_{\epsilon}$

-   We can use MSR to compute **standardized residuals**

    $$
    std.res_i = \frac{e_i}{\sqrt{MSR}}
    $$

-   Standardized residuals are produced by `augment()` in the column `.std.resid`
:::

## Using standardized residuals

We can examine the standardized residuals directly from the output from the `augment()` function

```{r}
#| echo: false
#| label: std residuals vs fitted
#| fig-align: center


ggplot(data = lemurs_aug, aes(x = .fitted, y = .std.resid)) + 
  geom_point() +
  geom_hline(yintercept = 3, color = "red", linetype =2)+
  geom_hline(yintercept = -3, color = "red", linetype = 2) +
  labs(x  = "Fitted values",
       y = "Std. Residuals",
       title = "Standardized residuals vs. fitted")
```

-   An observation is a *potential outlier* if its standardized residual is beyond $\pm 3$

## Studentized residuals {.midi}

::: incremental
-   MSR is an approximation of the variance of the residuals.

-   The variance of the residuals is $Var(\mathbf{e}) = \sigma^2_{\epsilon}(\mathbf{I} - \mathbf{H})$

    -   The variance of the $i^{th}$ residual is $Var(e_i) = \sigma^2_{\epsilon}(1 - h_{ii})$

-   The **studentized residual** is the residual rescaled by the more exact calculation for variance
:::

$$
r_i = \frac{e_{i}}{\sqrt{\hat{\sigma}^2_{\epsilon}(1 - h_{ii})}}
$$

-   Standardized and studentized residuals provide similar information about which points are outliers in the response.
    -   Studentized residuals are used to compute Cook's Distance.

## Using these measures

-   Standardized residuals, leverage, and Cook's Distance should all be examined together

-   Examine plots of the measures to identify observations that are outliers, high leverage, and may potentially impact the model.

## Back to the influential point

```{r}
#| echo: false
#| fig-align: center

lemurs_aug |>
  mutate(influential = factor(if_else(.cooksd > 1, 0,1))) |>
ggplot(aes(x = age_at_wt_mo, y = weight_g, color = influential)) + 
  geom_point(size = 2) +
  labs(x = "Age in months",
       y = "Weight in grams", 
       title = "Weight vs. Age of PCOQ lemurs") + 
   scale_color_viridis(discrete = T, end = 0.8) +
  theme(legend.position = "none")
```

## What to do with outliers/influential points?

::: incremental
-   First consider if the outlier is a result of a data entry error.

-   If not, you may consider dropping an observation if it's an outlier in the <u>predictor</u> variables if...

    -   It is meaningful to drop the observation given the context of the problem

    -   You intended to build a model on a smaller range of the predictor variables. Mention this in the write up of the results and be careful to avoid extrapolation when making predictions
:::

## What to do with outliers/influential points?

::: incremental
-   It is generally **not** good practice to drop observations that ar outliers in the value of the <u>response</u> variable

    -   These are legitimate observations and should be in the model

    -   You can try transformations or increasing the sample size by collecting more data

-   A general strategy when there are influential points is to fit the model with and without the influential points and compare the outcomes
:::

# Multicollinearity

## Data: Trail users {.midi}

-   The Pioneer Valley Planning Commission (PVPC) collected data at the beginning a trail in Florence, MA for ninety days from April 5, 2005 to November 15, 2005 to
-   Data collectors set up a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.

```{r}
#| echo: false
#| message: false
#| warning: false

rail_trail <- read_csv(here::here("slides", "data/rail_trail.csv"))
rail_trail |> slice(1:5)
```

Source: [Pioneer Valley Planning Commission](http://www.fvgreenway.org/pdfs/Northampton-Bikepath-Volume-Counts%20_05_LTA.pdf) via the **mosaicData** package.

## Variables

**Response**

-   `volume` estimated number of trail users that day (number of breaks recorded)

**Predictors**

-   `hightemp` daily high temperature (in degrees Fahrenheit)

-   `avgtemp` average of daily low and daily high temperature (in degrees Fahrenheit)

-   `season` one of “Fall”, “Spring”, or “Summer”

-   `precip` measure of precipitation (in inches)

## EDA: Relationship between predictors

We can create a pairwise plot matrix using the `ggpairs` function from the **GGally** R package

```{r}
#| eval: false

rail_trail |>
  select(hightemp, avgtemp, season, precip) |>
  ggpairs()
```

## EDA: Relationship between predictors {.midi}

```{r}
#| echo: false
#| fig-align: center

rail_trail |>
  select(hightemp, avgtemp, season, precip) |>
  ggpairs()
```

## EDA: Correlation matrix

We can. use `corrplot()` in the **corrplot** R package to make a matrix of pairwise correlations between [quantitative]{.underline} predictors

```{r}
#| eval: false
correlations <- rail_trail |>
  select(hightemp, avgtemp, precip) |>
  cor()

corrplot(correlations, method = "number")
```

## EDA: Correlation matrix {.midi}

```{r}
#| echo: false
#| fig-align: center

correlations <- rail_trail |>
  select(hightemp, avgtemp, precip) |>
  cor()


corrplot(correlations, method = "number")
```

::: question
What might be a potential concern with a model that uses high temperature, average temperature, season, and precipitation to predict volume?
:::

# Multicollinearity

## Multicollinearity

::: incremental
-   Ideally the predictors are orthogonal, meaning they are completely independent of one another

-   In practice, there is typically some dependence between predictors but it is often not a major issue in the model

-   If there is linear dependence among (a subset of) the predictors, we cannot find $\hat{\boldsymbol{\beta}}$

-   If there are near-linear dependencies, we can find $\hat{\boldsymbol{\beta}}$ but there may be other issues with the model

-   **Multicollinearity**: near-linear dependence among predictors
:::

## Sources of multicollinearity

::: incremental
-   Data collection method - only sample from a subspace of the region of predictors

-   Constraints in the population - e.g., predictors family income and size of house

-   Choice of model - e.g., adding high order terms to the model

-   Overdefined model - have more predictors than observations
:::

::: aside
Source: @montgomery2021introduction
:::

## Detecting multicollinearity

::: incremental
-   Recall $Var(\hat{\boldsymbol{\beta}}) = \sigma^2_{\epsilon}(\mathbf{X}^\mathsf{T}\mathbf{X})^{-1}$
-   Let $\mathbf{C} = (\mathbf{X}^\mathsf{T}\mathbf{X})^{-1}$. Then $Var(\hat{\beta}_j) = \sigma^2_{\epsilon}C_{jj}$
-   When there are near-linear dependencies, $C_{jj}$ increases and thus $Var(\hat{\beta}_j)$ becomes inflated
-   $C_{jj}$ is associated with how much $Var(\hat{\beta}_j)$ is inflated due to $x_j$ dependencies with other predictors
:::

## Variance inflation factor {.midi}

-   The **variance inflation factor (VIF)** measures how much the linear dependencies impact the variance of the predictors

$$
VIF_{j} = \frac{1}{1 - R^2_j}
$$

where $R^2_j$ is the proportion of variation in $x_j$ that is explained by a linear combination of all the other predictors

. . .

-   When the response and predictors are scaled in a particular way, $C_{jj} = VIF_{j}$. [Click here](variance-inflation-factors.html) to see how.

## Detecting multicollinearity

-   Common practice uses threshold $VIF > 10$ as indication of concerning multicollinearity (some say VIF \> 5 is worth investigation)

-   Variables with similar values of VIF are typically the ones correlated with each other

-   Use the `vif()` function in the **rms** R package to calculate VIF

```{r}
library(rms)

trail_fit <- lm(volume ~ hightemp + avgtemp + precip, data = rail_trail)

vif(trail_fit)
```

## How multicollinearity impacts model {.midi}

::: incremental
-   Large variance for the model coefficients that are collinear

    -   Different combinations of coefficient estimates produce equally good model fits

-   Unreliable statistical inference results

    -   May conclude coefficients are not statistically significant when there is, in fact, a relationship between the predictors and response

-   Interpretation of coefficient is no longer "holding all other variables constant", since this would be impossible for correlated predictors
:::

# Application exercise

::: appex
📋 [sta221-sp26.github.io/ae/ae-04-multicollinearity.html](../ae/ae-04-multicollinearity.html)

Part 1
:::

## Dealing with multicollinearity

::: incremental
-   Collect more data (often not feasible given practical constraints)

-   Redefine the correlated predictors to keep the information from predictors but eliminate collinearity

    -   e.g., if $x_1, x_2, x_3$ are correlated, use a new variable $(x_1 + x_2) / x_3$ in the model

-   For categorical predictors, avoid using levels with very few observations as the baseline

-   Remove one of the correlated variables

    -   Be careful about substantially reducing predictive power of the model
:::

# Application exercise

::: appex
📋 [sta221-sp26.github.io/ae/ae-04-multicollinearity.html](../ae/ae-04-multicollinearity.html)

Part 2
:::

## Recap

-   Reviewed model conditions and diagnostics

-   Introduced multicollinearity and how it impacts a model

## Next class

-   Variable transformations

-   Complete [Lecture 13 prepare](../prepare/prepare-lec13.html)

## References
