---
title: "Ridge regression"
author: "Prof. Maria Tackett"
date: "03-19-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 17-ridge-regression-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen=999)
```

## Announcements

-   HW 03 due March 19 at 11:59pm

-   Project exploratory data analysis due March 19

-   Statistics experience due April 2

-   DataFest March 20 - 22: [[https://dukestatsci.github.io/datafest/]{.underline}](https://dukestatsci.github.io/datafest/ "https://dukestatsci.github.io/datafest/")

## Topics

-   Ridge regression

## Computing set up

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)  
library(tidymodels)  
library(knitr)       
library(patchwork)
library(rms)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## AIC / BIC recap

## AIC and BIC in R

# Ridge regression

## Data: Age of abalones

The data for this analysis contains measurements for abalones, a type of marine snail. These measurements were collected and analyzed by researchers in Warwick et al. ([1994](https://sta221-sp26.github.io/hw/hw-03.html#ref-warwick1994population)). [Click here](https://www.researchgate.net/profile/Warwick-Nash/publication/287546509_7he_Population_Biology_of_Abalone_Haliotis_species_in_Tasmania_I_Blacklip_Abalone_H_rubra_from_the_North_Coast_and_Islands_of_Bass_Strait/links/5d949460458515202b7bf592/7he-Population-Biology-of-Abalone-Haliotis-species-in-Tasmania-I-Blacklip-Abalone-H-rubra-from-the-North-Coast-and-Islands-of-Bass-Strait.pdf) for the publication.

The 4177 abalones in this study can be reasonably treated as a random sample.

**The goal of this analysis is to use characteristics of the abalones to predict their age.**

```{r}
#| echo: false
abalones <- read_csv("data/abalone.csv")
```

## Variables

-   `Sex`: Male (M), Female (F), Infant (I)
-   `Length`: Longest shell measurement (in millimeters)
-   `Diameter`: Measured perpendicular to length (in millimeters)
-   `Height` : Measured with meat in shell (in millimeters)
-   `Whole_Weight`: Total weight of abalone (in grams)
-   `Age`: Age (in years)

## Fit a model  {.midi}

```{r}
age_fit <- lm(Age ~ Sex + Length + Diameter + Height
              + Whole_Weight, data = abalones)

tidy(age_fit) |> kable(digits = 3)
```

## Potential issue

```{r}
vif(age_fit)
```

<br>

::: question
-   What is the potential issue with this model?

-   How does this impact the model results?
:::

## Limitations of the least-squares estimator

-   When $(\mathbf{X}^\mathsf{T}\mathbf{X})$ is nearly singular, $Var(\hat{\boldsymbol{\beta}})$ gets large, and thus we have unstable coefficient estimates . This occurs when

    -   There are collinear predictors
    -   $p$ is large relative to $n$

-   Additionally, if $p > n$ , then we cannot find $\hat{\boldsymbol{\beta}}$

-   We can use a different estimator that does not have these limitations

## Finding the ridge estimator

We find the ridge estimator, $\hat{\boldsymbol{\beta}}_{ridge}$ by minimizing

$$
\underbrace{(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})^\mathsf{T}(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})}_{\text{SSR}} + \underbrace{\lambda \boldsymbol{\beta}^\mathsf{T}\boldsymbol{\beta}}_{\text{penalty}}
$$

where $\lambda \geq 0$ is the *tuning parameter* determined by the statistician (more on this later)

-   Like least-squares, ridge regression seeks to minimize the sum of squares

-   The penalty, $\lambda\boldsymbol{\beta}^\mathsf{T}\boldsymbol{\beta} = \lambda\sum_{j = 1}^p\beta_j$ preferences smaller values of $\beta_j$, so ridge will shrink the coefficients to 0 (this does not apply to the intercept)

## Finding the ridge estimator

We find the ridge estimator, $\hat{\boldsymbol{\beta}}_{ridge}$ by minimizing

$$
\underbrace{(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})^\mathsf{T}(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})}_{\text{SSR}} + \underbrace{\lambda \boldsymbol{\beta}^\mathsf{T}\boldsymbol{\beta}}_{\text{penalty}}
$$

<br>

::: question
What will happen to $\hat{\boldsymbol{\beta}}_{ridge}$ as $\lambda \rightarrow 0$? What will happen as$\lambda \rightarrow \infty$ ?
:::

## Set up problem

-   Abalone data set

-   Issue with multicollinearity

-   Variances will be high (show X\^TX matrix)

What if we want to keep all the predictors in the model?

## Ridge regression

-   show optimization

-   describe the penalty

-   what is happening as lambda goes to 0? What about as lambda goes to infinity? (We'll talk about how to find lambda later)

-   We will use standardized predictors for ridge regression. why does this matter?

## Estimator

-   you found this in exam 01

-   Is it an unbiased estimator?

-   what is the variance? - this may be too long for class?

-   This is the bias-variance trade off - we accept a little bias in order to reduce the variance

## When is ridge useful

Most useful for models in which the SE of OLS are large and unstable but you want to keep all the predictors

-   That lambda I term helps stablize the variance when x\^tX is nearly singular.

-   So this is the case when there is high multicollinearity

-   when P is close to n

-   Makes it possible to estimate coefficients when p \> n (is this true)?

## LImitations

-   Ridge will not ever set the coefficients to exactly equal to 0 (limitation if you want to reduce the number of coefficients in th emodel)

    -   Similar method called LASSO that addresses this

-   

## Further reading

Link to original paper (reprinted version) <https://www.jstor.org/stable/1271436?seq=6>

## data

```{r}
library(tidyverse)
library(tidymodels)
library(rms)

nba_salaries <- read_csv("data/nba-salaries-2022-23.csv")

nba_salaries <- nba_salaries |>
  mutate(PG = if_else(Position == "PG", TRUE, FALSE))

nba_salaries <- nba_salaries |>
  select(-Position,  -`Player Name`) 

nba_salaries_scaled <- nba_salaries |>
  mutate(across(-Salary, ~ if(is.numeric(.)) scale(.) else .))

model1 <- lm(Salary ~ ., data = nba_salaries)
tidy(model1)

vif(model1)
```

```{r}
#| label: ridge-model 

# pass standardized data in: 
ridge_fit <- MASS::lm.ridge(Salary ~ 0 + ., data = nba_salaries_scaled,
                           lambda = 5000)

beta_scaled_lmridge <-  ridge_fit$coef # scaled space coefficients
beta_unscaled_lmridge <-  coef(ridge_fit) # original scale coefficients 

# (different than computed above if y and X are not standardized inputs)

tidy(ridge_fit)

```

## Questions

-   Form for questions to discuss in lab.

## Next class

-   Ridge regression

-   No prepare assignment
