---
title: "Ridge regression"
author: "Prof. Maria Tackett"
date: "03-19-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 17-ridge-regression-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen=999)
```

## Announcements

-   HW 03 due March 19 at 11:59pm

-   Project exploratory data analysis due March 19

-   Statistics experience due April 2

-   DataFest March 20 - 22: [[https://dukestatsci.github.io/datafest/]{.underline}](https://dukestatsci.github.io/datafest/ "https://dukestatsci.github.io/datafest/")

## Topics

-   Ridge regression

## AIC / BIC recap

## AIC and BIC in R

# Ridge regression 

## Set up problem 

-   Abalone data set

-   Issue with multicollinearity

-   Variances will be high (show X\^TX matrix)

What if we want to keep all the predictors in the model?

## Ridge regression 

-   show optimization

-   describe the penalty

-   what is happening as lambda goes to 0? What about as lambda goes to infinity? (We'll talk about how to find lambda later)

-   We will use standardized predictors for ridge regression. why does this matter?

## Estimator 

-   you found this in exam 01

-   Is it an unbiased estimator?

-   what is the variance? - this may be too long for class?

-   This is the bias-variance trade off - we accept a little bias in order to reduce the variance

## When is ridge useful 

Most useful for models in which the SE of OLS are large and unstable but you want to keep all the predictors

-   That lambda I term helps stablize the variance when x\^tX is nearly singular.

-   So this is the case when there is high multicollinearity

-   when P is close to n

-   Makes it possible to estimate coefficients when p \> n (is this true)?

## LImitations 

-   Ridge will not ever set the coefficients to exactly equal to 0 (limitation if you want to reduce the number of coefficients in th emodel)

    -   Similar method called LASSO that addresses this

-   

## Further reading 

Link to original paper (reprinted version) <https://www.jstor.org/stable/1271436?seq=6>

## data

```{r}
library(tidyverse)
library(tidymodels)
library(rms)

nba_salaries <- read_csv("data/nba-salaries-2022-23.csv")

nba_salaries <- nba_salaries |>
  mutate(PG = if_else(Position == "PG", TRUE, FALSE))

nba_salaries <- nba_salaries |>
  select(-Position,  -`Player Name`) 

nba_salaries_scaled <- nba_salaries |>
  mutate(across(-Salary, ~ if(is.numeric(.)) scale(.) else .))

model1 <- lm(Salary ~ ., data = nba_salaries)
tidy(model1)

vif(model1)
```

```{r}
#| label: ridge-model 

# pass standardized data in: 
ridge_fit <- MASS::lm.ridge(Salary ~ 0 + ., data = nba_salaries_scaled,
                           lambda = 5000)

beta_scaled_lmridge <-  ridge_fit$coef # scaled space coefficients
beta_unscaled_lmridge <-  coef(ridge_fit) # original scale coefficients 

# (different than computed above if y and X are not standardized inputs)

tidy(ridge_fit)

```

## Questions 

-   Form for questions to discuss in lab.

## Next class

-   Ridge regression

-   No prepare assignment
