---
title: "Logistic Regression"
subtitle: "Model selection"
author: "Prof. Maria Tackett"
date: "03-31-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 20-logistic-selection-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

## Announcements {.midi}

-   Presentation comments due TODAY at 11:59pm

-   Statistics experience due April 2 at 11:59pm

-   Next project milestone: Draft report due April 10 (before lab)

# Presentations

# Model selection

## Topics

-   Comparing models using AIC and BIC


## Computational setup

```{r}
#| echo: true
#| warning: false
#| message: false


library(tidyverse)
library(tidymodels)
library(pROC)      
library(knitr)
library(kableExtra)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `totChol`: Total cholesterol (in mg/dL)

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

```{r}
#| echo: false
#| message: false
#| warning: false


heart_disease <- read_csv("data/framingham.csv") |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |>
  mutate(
    high_risk = as_factor(TenYearCHD),
    education = as_factor(education),
    currentSmoker = as_factor(currentSmoker)
  )
```

## Modeling risk of coronary heart disease

Using `age`, `totChol`, and `currentSmoker`

```{r}
#| echo: false
high_risk_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

high_risk_fit2 <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")

tidy(high_risk_fit, conf.int = TRUE) |> 
  kable(format = "markdown", digits = 3)
```

## Review: ROC Curve + Model fit

```{r}
#| echo: false
high_risk_aug <- augment(high_risk_fit)

roc_curve_data <- high_risk_aug |>
  roc_curve(high_risk, .fitted, event_level = "second")

autoplot(roc_curve_data)

#calculate AUC
high_risk_aug |>
  roc_auc(high_risk, .fitted, event_level = "second")
```

## Review: Classification

We will use a threshold of 0.2 to classify observations

```{r}
#| echo: false
#compute predicted probabilities
pred_prob <- predict.glm(high_risk_fit, type = "response")
high_risk_aug <- high_risk_aug |> 
  bind_cols(pred_prob = pred_prob) |>
  mutate(pred_class = factor(if_else(pred_prob > 0.2, 1, 0)))

# compute confusion matrix
high_risk_aug |>
  conf_mat(high_risk, pred_class) |>
  autoplot(type = "heatmap")
```

## Review: Classification {.midi}

```{r}
#| echo: false

high_risk_aug |>
  conf_mat(high_risk, pred_class) |>
  autoplot(type = "heatmap")
```

::: question
1.  Compute the misclassification rate.

2.  Compute sensitivity and explain what it means in the context of the data.

3.  Compute specificity and explain what it means in the context of the data.
:::

# Model comparison

## Which model do we choose?

```{r}
#| echo: false

high_risk_fit2 <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")
```

::::: columns
::: {.column width="50%"}
<center>**Model 1**</center>

```{r}
#| echo: false

tidy(high_risk_fit) |> 
  select(term, estimate) |>
  kable(digits = 3)

```
:::

::: {.column width="50%"}
<center>**Model 2**</center>

```{r}
#| echo: false

tidy(high_risk_fit2) |> 
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

## Log-Likelihood

Recall the log-likelihood function

$$
\begin{aligned}
\log L&(\boldsymbol{\beta}|x_1, \ldots, x_n, y_1, \dots, y_n) \\
&= \sum\limits_{i=1}^n[y_i \log(\pi_i) + (1 - y_i)\log(1 - \pi_i)]
\end{aligned}
$$

where $\pi_i = \frac{\exp\{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}{1 + \exp\{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}$

## AIC & BIC

Estimators of prediction error and *relative* quality of models:

. . .

**Akaike's Information Criterion (AIC)**[^1]: $$AIC = -2\log L + 2 (p+1)$$

[^1]: Akaike, Hirotugu. ["A new look at the statistical model identification."](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=1100705) *IEEE transactions on automatic control* 19.6 (1974): 716-723.

. . .

**Schwarz's Bayesian Information Criterion (BIC)**[^2]: $$ BIC = -2\log L + \log(n)\times(p+1)$$

[^2]: Schwarz, Gideon. ["Estimating the dimension of a model."](https://projecteuclid.org/journalArticle/Download?urlId=10.1214%2Faos%2F1176344136) *The annals of statistics* (1978): 461-464.

## AIC & BIC

$$
\begin{aligned} 
& AIC = \color{blue}{-2\log L}  \color{black}{+ 2(p+1)} \\
& BIC = \color{blue}{-2\log L}  + \color{black}{\log(n)\times(p+1)}
\end{aligned}
$$

. . .

<br>

First Term: Decreases as *p* increases

## AIC & BIC

$$
\begin{aligned} & AIC = -2\log L  + \color{blue}{2(p+1)} \\
& BIC = -2\log L + \color{blue}{\log(n)\times (p+1)} 
\end{aligned}
$$

<br>

Second term: Increases as *p* increases

## Using AIC & BIC

$$
\begin{aligned} & AIC = -2\log L  + \color{red}{2(p+1)} \\
& BIC = -2 \log L  + \color{red}{\log(n)\times(p+1)} 
\end{aligned}
$$

-   Choose model with the smaller value of AIC or BIC

-   If $n \geq 8$, the **penalty** for BIC is larger than that of AIC, so BIC tends to favor *more parsimonious* models (i.e. models with fewer terms)

## AIC from the `glance()` function

Let's look at the AIC for the model that includes `age`, `totChol`, and `currentSmoker`

```{r}
glance(high_risk_fit)$AIC
```

<br>

. . .

**Calculating AIC**

```{r}
- 2 * glance(high_risk_fit)$logLik + 2 * (3 + 1)
```

## Comparing the models using AIC

Let's compare the full and reduced models using AIC.

```{r}
#| echo: false

# AIC reduced model 
high_risk_fit_reduced <- high_risk_fit

# AIC full model
high_risk_fit_full <- high_risk_fit2

```

```{r}
# AIC reduced model 
glance(high_risk_fit_reduced)$AIC

# AIC full model
glance(high_risk_fit_full)$AIC
```

<br>

::: question
Based on AIC, which model do you select?
:::

## Comparing the models using BIC

Let's compare the full and reduced models using BIC

```{r}
#| echo: true

# BIC reduced model
glance(high_risk_fit_reduced)$BIC

# BIC full model
glance(high_risk_fit_full)$BIC
```

<br>

::: question
Based on BIC, which model do you select?
:::


## Recap

-   Introduced model selection for logistic regression using

    -   AIC and BIC

## Next class

-   Inference for logistic regression

-   Complete [Lecture 21 prepare](../prepare/prepare-lec21.html)

## References
