---
title: "Logistic Regression"
subtitle: "Inference"
author: "Prof. Maria Tackett"
date: "03-31-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 21-logistic-inference-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

## Announcements

-   Statistics experience due TODAY at 11:59pm

-   HW 04 due Thursday, April 9 at 11:59pm

    -   released after class

-   Next project milestone: Draft report due Friday, April 10 (before lab)

## Topics

-   Drop-in-deviance test

-   Inference for an individual coefficient

## Computational setup

```{r}
#| echo: true
#| warning: false
#| message: false


library(tidyverse)
library(tidymodels)
library(pROC)      
library(knitr)
library(kableExtra)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `totChol`: Total cholesterol (in mg/dL)

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

```{r}
#| echo: false
#| message: false
#| warning: false


heart_disease <- read_csv("data/framingham.csv") |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |>
  mutate(
    high_risk = as_factor(TenYearCHD),
    education = as_factor(education),
    currentSmoker = as_factor(currentSmoker)
  )
```

## Which model do we choose?

```{r}
#| echo: false

high_risk_fit <-  glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

high_risk_fit_reduced <- high_risk_fit

high_risk_fit2 <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")

high_risk_fit_full <- high_risk_fit2
```

::::: columns
::: {.column width="50%"}
<center>**Model 1**</center>

```{r}
#| echo: false

tidy(high_risk_fit) |> 
  select(term, estimate) |>
  kable(digits = 3)

```
:::

::: {.column width="50%"}
<center>**Model 2**</center>

```{r}
#| echo: false

tidy(high_risk_fit2) |> 
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

## Log-Likelihood

Recall the log-likelihood function

$$
\begin{aligned}
\log L&(\boldsymbol{\beta}|x_1, \ldots, x_n, y_1, \dots, y_n) \\
&= \sum\limits_{i=1}^n[y_i \log(\pi_i) + (1 - y_i)\log(1 - \pi_i)]
\end{aligned}
$$

where $\pi_i = \frac{\exp\{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}{1 + \exp\{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}$

## Comparing the models using AIC and BIC

::::: columns
::: {.column width="50%"}
**AIC**

```{r}
# AIC reduced model 
glance(high_risk_fit_reduced)$AIC

# AIC full model
glance(high_risk_fit_full)$AIC
```
:::

::: {.column width="50%"}
**BIC**

```{r}

# BIC reduced model
glance(high_risk_fit_reduced)$BIC

# BIC full model
glance(high_risk_fit_full)$BIC
```
:::
:::::

# Drop-in-deviance test

## Drop-in-deviance test

We can use the **drop-in-deviance test** (aka Likelihood Ratio Test) to test

-   the overall statistical significance of a logistic regression model

-   the statistical significance of a subset of coefficients in the model

## Deviance {.midi}

The **deviance** is a measure of the degree to which the predicted values are different from the observed values (compares the current model to a "saturated" model)

In logistic regression,

$$
D = -2 \log L 
$$

<br>

$D \sim \chi^2_{n - p - 1}$ ($D$ follows a Chi-square distribution with $n - p - 1$ degrees of freedom)[^1]

[^1]: See @wilks1935likelihood for explanation of why -2 is included.

<br>

Note: $n - p - 1$ a the degrees of freedom associated with the error in the model (like residuals)

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Test for overall significance

We can test the overall significance for a logistic regression model, i.e., whether there is at least one predictor with a non-zero coefficient

$$
\begin{aligned}
&H_0: \beta_1 = \dots = \beta_p = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one } j
\end{aligned}
$$

. . .

The **drop-in-deviance test for overall significance** compares the fit of a model with no predictors to the current model.

## Drop-in-deviance test statistic {.midi}

Let $L_0$ and $L_a$ be the likelihood functions of the model under $H_0$ and $H_a$, respectively. The **test statistic** is

$$
\begin{aligned}
G = D_0 - D_a &= (-2\log L_0) - (-2\log L_a)\\[5pt] 
& = -2(\log L_0 - \log L_a) \\[5pt]
&= -2\sum_{i=1}^n \Big[ y_i \log \Big(\frac{\hat{\pi}^0}{\hat{\pi}^a_i}\Big) + (1 - y_i)\log \Big(\frac{1-\hat{\pi}^0}{1-\hat{\pi}^a_i}\Big)\Big]
\end{aligned}
$$

where $\hat{\pi}^0$ is the predicted probability under $H_0$ and $\hat{\pi}_i^a = \frac{\exp \{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}{1 + \exp \{\mathbf{x}_i^\mathsf{T}\boldsymbol{\beta}\}}$ is the predicted probability under $H_a$ [^2]

[^2]: See @wilks1935likelihood for explanation of why -2 is included.

## Drop-in-deviance test statistic

$$
G = -2\sum_{i=1}^n \Big[ y_i \log \Big(\frac{\hat{\pi}^0}{\hat{\pi}^a_i}\Big) + (1 - y_i)\log \Big(\frac{1-\hat{\pi}^0}{1-\hat{\pi}^a_i}\Big)\Big]
$$

. . .

::: incremental
-   When $n$ is large, $G \sim \chi^2_p$, ( $G$ follows a Chi-square distribution with $p$ degrees of freedom)

-   The p-value is calculated as $P(\chi^2 > G)$

-   Large values of $G$ (small p-values) indicate at least one $\beta_j$ is non-zero
:::

## Heart disease model: drop-in-deviance test

$$
\begin{aligned}
&H_0: \beta_{age} = \beta_{totChol} = \beta_{currentSmoker} = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}$$

. . .

**Fit the null model** (we've already fit the alternative model)

```{r}
#| echo: true

null_model <- glm(high_risk ~ 1, data = heart_disease, family = "binomial")
```

```{r}
#| echo: false

tidy(null_model) |>
  kable()
```

## Heart disease model: drop-in-deviance test

**Compute the log-likelihoods**

```{r}
(L_0 <- glance(null_model)$logLik)
(L_a <- glance(high_risk_fit)$logLik)
```

. . .

**Compute the test statistic**

```{r}
(G <- -2 * (L_0 - L_a))
```

. . .

## Heart disease model: drop-in-deviance test

**Compute the p-value**

```{r}
pchisq(G, df = 3, lower.tail = FALSE)
```

. . .

**Conclusion**

The p-value is small, so we reject $H_0$. The data provide evidence that at least one predictor in the model has a non-zero coefficient.

## Why use overall test? {.midi}

Why do we use a test for overall significance instead of just looking at the test for individual coefficients?[^3]

[^3]: Example from *Introduction to Statistical Learning*

. . .

Suppose we have a model such that $p = 100$ and $H_0: \beta_1 = \dots = \beta_{100} = 0$ is true

. . .

::: incremental
-   About 5% of the p-values for individual coefficients will be below 0.05 by chance.

-   So we expect to see 5 small p-values if even no linear association actually exists.

-   Therefore, it is very likely we will see at least one small p-value by chance.

-   The overall test of significance does not have this problem. There is only a 5% chance we will get a p-value below 0.05, if a relationship truly does not exist.
:::

# Test a subset of coefficients

## Testing a subset of coefficients

-   Suppose there are two models:

    -   **Reduced Model**: includes predictors $x_1, \ldots, x_q$

    -   **Full Model**: includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

-   We can use a **drop-in-deviance test** to determine if any of the new predictors are useful

. . .

$$
\begin{aligned}
&H_0: \beta_{q+1} = \dots = \beta_p = 0\\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

## Drop-in-deviance test {.midi}

$$
\begin{aligned}
&H_0: \beta_{q+1} = \dots = \beta_p = 0\\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

. . .

The test statistic is

$$
\begin{aligned}
G = D_{reduced} - D_{full} &= (-2\log L_{reduced}) - (-2 \log L_{full}) \\
&= -2(\log L_{reduced} - \log L_{full})
\end{aligned}
$$

. . .

The p-value is calculated using a $\chi_{\Delta df}^2$ distribution, where $\Delta df$ is the number of parameters being tested (the difference in number of parameters between the full and reduced model).

## Example: Include `education`?

Should we include `education` in the model?

-   Reduced model: `age`, `totChol`, `currentSmoker`

-   Full model: `age`, `totChol`, `currentSmoker` , `education`

. . .

$$
\begin{aligned}
&H_0: \beta_{ed2} = \beta_{ed3} = \beta_{ed4} = 0 \\
&H_a: \beta_j \neq 0 \text{ for at least one }j
\end{aligned}
$$

## Example: Include `education`?

```{r}
reduced_model <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

full_model <- glm(high_risk ~ age + totChol + currentSmoker + education, 
              data = heart_disease, family = "binomial")
```

. . .

**Compute deviances**

```{r}
(deviance_reduced <- -2 * glance(reduced_model)$logLik)
(deviance_full <- -2 * glance(full_model)$logLik)
```

. . .

**Compute test statistic**

```{r}
(G <- deviance_reduced - deviance_full)
```

## Example: Include `education`?

**Compute p-value**

```{r}
pchisq(G, df = 3, lower.tail = FALSE)
```

<br>

. . .

::: question
What is your conclusion? Would you include `education` in the model that already has `age`, `totChol`, `currentSmoker`?
:::

## Drop-in-deviance test in R {.midi}

Conduct the drop-in-deviance test using the `anova()` function in R with option `test = "Chisq"`

```{r}
anova(reduced_model, full_model, test = "Chisq") |> 
  tidy() |> 
  kable(digits = 3)
```

## Add interactions with `currentSmoker`? {.midi}

```{r}
#| echo: false

new_model <- glm(high_risk ~ age + totChol + currentSmoker + currentSmoker*age + currentSmoker * totChol, data = heart_disease, family = "binomial")

anova(reduced_model, new_model, test = "Chisq") |>
  tidy() |>
  kable(digits = 3)

```

# Inference for an individual coefficient

## Distribution of $\hat{\boldsymbol{\beta}}$

When $n$ is large, $\hat{\boldsymbol{\beta}}$ is normally distributed

<br>

::: question
How do we know $\hat{\boldsymbol{\beta}}$ is normally for large $n$?
:::

## Distribution of $\hat{\boldsymbol{\beta}}$

When $n$ is large...

The expected value of $\hat{\boldsymbol{\beta}}$ is the true parameter, $\boldsymbol{\beta}$, i.e., $E(\hat{\boldsymbol{\beta}}) = \boldsymbol{\beta}$

. . .

$Var(\hat{\boldsymbol{\beta}})$, the matrix of variances and covariances between estimators

$$
Var(\hat{\boldsymbol{\beta}}) = (\mathbf{X}^\mathsf{T}\mathbf{V}\mathbf{X})^{-1}
$$

where $\mathbf{V}$ is a $n\times n$ diagonal matrix, such that $V_{ii}$ is the estimated variance for the $i^{th}$ observation

## Test for a single coefficient

**Hypotheses:** $H_0: \beta_j = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_j \neq 0$, given the other variables in the model

. . .

**(Wald) Test Statistic:** $$z = \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

where $SE(\hat{\beta}_j)$ is the square root of the $j^{th}$ diagonal element of $Var(\hat{\boldsymbol{\beta}})$

. . .

**P-value:** $P(|Z| > |z|)$, where $Z \sim N(0, 1)$, the Standard Normal distribution

## Confidence interval for $\beta_j$

We can calculate the **C% confidence interval** for $\beta_j$ as the following:

$$
\Large{\hat{\beta}_j \pm z^* \times SE(\hat{\beta}_j)}
$$

where $z^*$ is calculated from the $N(0,1)$ distribution

. . .

::: callout-note
This is an interval for the change in the log-odds for every one unit increase in $x_j$
:::

## Interpretation in terms of the odds

The change in **odds** for every one unit increase in $x_j$.

$$
\Large{\exp\{\hat{\beta}_j \pm z^* \times SE(\hat{\beta}_j)\}}
$$

. . .

**Interpretation:** We are $C\%$ confident that for every one unit increase in $x_j$, the odds multiply by a factor of $\exp\{\hat{\beta}_j - z^* \times SE(\hat{\beta}_j)\}$ to $\exp\{\hat{\beta}_j + z^* \times SE(\hat{\beta}_j)\}$, holding all else constant.

## Coefficient for `age` {.midi}

```{r}
#| label: risk-model-age-highlight
#| echo: false

heart_disease_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")

tidy(heart_disease_fit, conf.int = TRUE) |> 
  kable(digits = 3) |>
  row_spec(2, background = "#D9E3E4")
```

. . .

**Hypotheses:**

$$
H_0: \beta_{age} = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_{age} \neq 0
$$, given total cholesterol and smoking status are in the model.

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Test statistic:**

$$z = \frac{ 0.0825  - 0}{0.00575} = 14.34
$$

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**P-value:**

$$P(|Z| > |14.34|) \approx 0
$$

. . .

```{r}
2 * pnorm(14.34,lower.tail = FALSE)
```

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Conclusion:**

The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that age is a statistically significant predictor of whether someone is high risk of having heart disease, after accounting for total cholesterol and smoking status.

## CI for `age`

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

::: question
Interpret the 95% confidence interval for `age` in terms of the **odds** of being high risk for heart disease.
:::

## Let's take a look at `education`

```{r}
#| echo: false
 
tidy(high_risk_fit_full, conf.int = TRUE) |> 
  kable(digits = 3) 
```

::: question
What is your final assessment about including `education` in the model?
:::

## Overview of testing coefficients

**Test a single coefficient**

-   Drop-in-deviance test

-   Wald hypothesis test and confidence interval

. . .

**Test a subset of coefficients**

-   Drop-in-deviance test

. . .

Can use AIC and BIC to compare models in both scenarios

## Recap

-   Model selection for logistic regression using a drop-in-deviance test

-   Inference for an individual coefficient

## Next class

-   Data science ethics

-   No prepare assignment

## References
