---
title: "Final exam review"
author: "Prof. Maria Tackett"
date: "04-21-2026"
date-format: "MMM DD, YYYY"
footer: "[🔗 sta221-sp26.github.io](https://sta221-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 24-final-exam-review-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
library(tidymodels)
library(knitr)
```

## Announcements

-   Project report and organized GitHub repo - due Wednesday, April 22 at 11:59pm

-   Course and TA evaluations - open until April 25 at 11:59pm

-   Final exam: April 30, 9am - 12pm

## Project

Any questions about the project?

## Course evaluations

Please share your feedback about the course!

Course evaluations [**and**]{.underline} TA evaluations are open now until April 25 at 11:59pm.

-   If the response rate is at least 80% on both course evaluations [**and**]{.underline} TA evaluations, everyone in the class will receive 0.5 points (out of 50) added to their final exam grade.

-   Check email for links to course and TA evaluations.

## Final exam

-   In-class only: April 30, 9am - 12pm

-   Cumulative exam covering content from the full semester

-   Format similar to Exam 01 and Exam 02

-   Official university documentation or dean's excuse required to excuse the exam

-   Can bring one page (front and back) of handwritten notes written directly on paper

    -   Turn in with the exam

    -   Point deductions for notes that don't follow rules

## Topics {.midi}

::::: columns
::: {.column width="50%"}
**Linear regression**

-   Fitting and interpreting models
-   Model evaluation
-   Different types of predictors
-   Inference
-   Matrix representation of regression
-   Model conditions and diagnostics
-   Multicollinearity
-   Variable transformations
:::

::: {.column width="50%"}
-   Maximum likelihood estimation
-   Properties of estimators
-   Model selection
-   Ridge regression

**Logistic regression**

-   Fitting and interpreting models
-   Predicted probabilities and classes
-   ROC curve and AUC
-   Inference
-   Model selection

**Data science ethics**
:::
:::::

## Tips for studying {.midi}

-   Rework derivations from assignments and lecture notes

-   Review in-class activities and assignments, asking "why" as you review your process and reasoning

-   Focus on understanding not memorization

-   Explain concepts / process to others

-   Ask questions in office hours

-   Review prepare readings

-   Review lecture recordings (available until start of final exam)

## Resources

-   Prepare readings

-   Lecture notes

-   Lecture recordings available until start of exam

-   HW and lab assignments (all solutions on Canvas)

-   Exam 01 and Exam 02 practice problems

# Practice

## Logistic regression response variable

$Y$ is a binary response variable.

-   What is the distribution of $Y$?

-   What is the variance of $y_i$, given the logistic regression model?

-   Why doesn't constant variance hold for logistic regression?

## Logistic regression coefficients

Let $\pi_i$ be the probability an individual is diagnosed with lung cancer. We fit the model

$$
\log\Big(\frac{\hat{\pi}_i}{1-\hat{\pi}_i}\Big) = \hat{\beta}_0 + \hat{\beta}_1 \text{birdkeepingYes}
$$

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false

df <- read_csv("data/birdkeeping.csv") |>
  mutate(cancer = factor(cancer), 
         bird = factor(if_else(bird == 1, "Yes", "No")), 
         sex = factor(if_else(female == 1, "Female", "Male")), 
         socioeconomic = factor(if_else(highstatus == 1, "High", "Low"))
  )

bird_model <- glm(cancer ~ bird, data = df, family = "binomial")

tidy(bird_model) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
-   Explain what -1.386 tells us.

-   Explain what 1.356 tells us.
:::
:::::

## Logistic regression odds ratios

```{r}
#| echo: false

# birdkeeping data

df |>
  mutate(cancer_names = if_else(cancer == "1", "Cancer", "No Cancer")) |>
  count(bird, cancer_names) |>
  pivot_wider(names_from = cancer_names, values_from = n) |>
  kable(col.names = c("Birdkeeping", "Cancer", "No Cancer"))
```

## Drop-in-deviance test

```{r}
#| echo: false
model1 <- glm(cancer ~ socioeconomic + age + bird , data = df,
              family = "binomial")


model1_full <- glm(cancer ~ socioeconomic + age + bird +
                     yrsmoke + cigsday, data = df,
              family = "binomial")

anova(model1, model1_full, test = "Chisq") |> 
  tidy() |>
  kable(digits = 0, col.names = c("term", "df.resid", "deviance", "df", "G", "p.value"))
```

-   What is the distribution of the test statistic $G$?

-   How would do we know the p-value will be very small?

## Unbiased estimator

-   What does it mean for an estimator to be unbiased?

-   How do you show an estimator is unbiased?

## Linear regression

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

-   What assumptions on $\boldsymbol{\epsilon}$ are needed to derive the least-squares estimator?

-   What assumptions on $\boldsymbol{\epsilon}$ are needed to conduct inference for $\boldsymbol{\beta}$ ?

-   What assumptions on $\boldsymbol{\epsilon}$ are needed to derive the MLE for $\boldsymbol{\beta}$?

## Sum of squares

$$
\text{Show }\hspace{1mm} SSR = \mathbf{y}^\mathsf{T}\mathbf{y} - \hat{\boldsymbol{\beta}}^\mathsf{T}\mathbf{X}^\mathsf{T}\mathbf{y}
$$

## Types of predictors

-   Why might we use centered predictors in a model?

-   Why might we use standardized predictors in a model?

# Questions?
