---
title: "HW 01: Simple linear regression"
subtitle: Lemurs
bibliography: references.bib
---

::: callout-important
## Due date

This assignment is due on **Thursday, January 29 at 11:59pm**. To be considered on time, the following must be done by the due date:

-   Final `.qmd` and `.pdf` files pushed to your GitHub repo
-   Final `.pdf` file submitted on Gradescope
:::

# Introduction 

You will use simple linear regression to explore the growth rate of young lemurs in the Duke Lemur Center. You will also explore the mathematical properties of simple linear regression models.

# Learning goals

In this assignment, you will…

-   use matrix operations to show results about simple linear regression.
-   conduct exploratory data analysis.
-   fit and interpret simple linear regression models.
-   evaluate model fit.
-   continue developing a workflow for reproducible data analysis.

# Getting started

-   Go to the [sta221-sp26](https://www.github.com/sta221-sp26) organization on GitHub. Click on the repo with the prefix **hw-01**. It contains the starter documents you need to complete the lab.

-   Clone the repo and start a new project in RStudio. See the [Lab 00 instructions](../labs/lab-00.html) for details on cloning a repo and starting a new project in R.

# Packages

The following packages are used in this assignment:

```{r}
#| eval: false

library(tidyverse)
library(tidymodels)
library(knitr)

# load other packages as needed
```

# Conceptual exercises

## Instructions

The conceptual exercises are focused on explaining concepts and showing results mathematically. Show your work for each question.\
\
**You may write the answers and associated work for conceptual exercises by hand or type them in your Quarto document.**

## Exercise 1

In lecture, we introduced $\mathbf{H}$, the hat matrix, as a projection matrix that projects $\mathbf{y}$ onto $\text{Col}(\mathbf{X})$ . Here we will show some properties of $\mathbf{H}$.

a\. Show that the hat matrix $\mathbf{H}$ is symmetric $(\mathbf{H}^\mathsf{T} = \mathbf{H})$ and idempotent $(\mathbf{H}^2 = \mathbf{H})$.

d\. Show that all eigenvalues of $\mathbf{H}$ are 0 or 1.

## Exercise 2

*This exercise is adapted from @casella2024statistical.*

Suppose there are $n$ observations $(x_1, y_1), \ldots, (x_n, y_n)$, such that the relationship between $X$ and $Y$ can be summarized as $$
y_i = \beta x_i^2 + \epsilon_i \hspace{8mm} \epsilon_i \sim N(0,\sigma^2_{\epsilon})
$$

Find, $\hat{\beta}$, the least-squares estimator of $\beta$.

## Exercise 3

In class we used the sum of squared errors, $\boldsymbol{\epsilon}^\mathsf{T}\boldsymbol{\epsilon}$ , to estimate the regression coefficients, $\hat{\boldsymbol{\beta}} = (\mathbf{X}^\mathsf{T}\mathbf{X})^{-1}\mathbf{X}^\mathsf{T}\mathbf{Y}$ . To show this is the least squares estimator, we now need to show that we have, in fact, found the estimator for $\boldsymbol{\beta}$ that minimizes the sum of square errors.

If the Hessian matrix $\nabla_{\boldsymbol{\beta}}^2 \boldsymbol{\epsilon}^\mathsf{T}\boldsymbol{\epsilon}$ is positive definite, then we know we have found the $\hat{\boldsymbol{\beta}}$ that minimizes the sum of squared errors, i.e., the least squares estimator.

Show that $\nabla_{\boldsymbol{\beta}}^2 \boldsymbol{\epsilon}^\mathsf{T}\boldsymbol{\epsilon} \propto \mathbf{X}^\mathsf{T}\mathbf{X}$ is positive definite.

## Exercise 4

*This exercise is adapted from @montgomery2021introduction .*

Prove that the maximum value of $R^2$ must be less than 1 if the data set contains observations such that there are different observed values of the response for the same value of the predictor (e.g., the data set contains observations $(x_i, y_i)$ and $(x_j, y_j)$ such that $x_i = x_j$ and $y_i \neq y_j$ ).

## Exercise 5

Show that the sum of squared residuals (SSR) can be written as the following:

$$
\mathbf{y}^\mathsf{T}\mathbf{y} - \hat{\boldsymbol{\beta}}^\mathsf{T}\mathbf{X}^\mathsf{T}\mathbf{y}
$$

# Applied exercises

## Instructions

The applied exercises are focused on applying the concepts to analyze data.

**All work for the applied exercises must be typed in your Quarto document following a reproducible workflow.**

Write all narrative using complete sentences and include informative axis labels / titles on visualizations.

## Data - UPDATE 

\[ADD DESCRIPTION\]

```{r}
#| eval: false

lemurs <- read_csv("https://intro-regression.github.io/data/lemurs-sample-young.csv")
```

The analysis will focus on the following variables:

-   `age`: Age of the lemur in months

-   `weight`: Weight of the lemur in grams

## Analysis goal

The goal of this analysis is to use linear regression explain variability in ice duration for lakes in Madison, WI based on air temperature. Because ice cover is impacted by various environmental factors, researchers are interested in examining the association between these two factors to better understand the changing climate.

## **Exercise 6**

Let's start by looking at the response and predictor variables.

a.  Visualize distribution of `weight`. Describe the distrbution.

b.  Visualize the distribution of `age`. Describe the distribution.

::: callout-tip
The description should include shape, center, spread, and presence of outliers. Use specific values in your description. See [Section 3.4.2](https://intro-regression.github.io/03-eda.html#sec-univar-eda-quant) of *Introduction to Regression Analysis* for more information about describing univariate distributions. See the [ggplot2 reference](https://ggplot2.tidyverse.org/reference/index.html#geoms) for example code and plots.
:::

## Exercise 7

Visualize the relationship between `weight` and `age`. Do you think a linear model is a reasonable choice to summarize the relationship between the two variables? Briefly explain.

::: callout-important
Now is a good time to render your document again if you haven't done so recently and commit (with a meaningful commit message) and push all updates.
:::

## Exercise 8

We will fit a model using `age` to explain variability in the `weight`. The model takes the form

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

a.  State the dimensions of $\mathbf{y}$, $\mathbf{X}$, $\boldsymbol{\beta}$, $\boldsymbol{\epsilon}$ for this analysis. Your answer should have exact values given this data set.
b.  Estimate the regression coefficients $\hat{\boldsymbol{\beta}}$ in R using the matrix representation. Show the code used to get the answer.
c.  Check your results from part (b) by using the `lm` function to fit the model. Neatly display your results using 3 digits.

## Exercise 9

a.  Calculate $R^2$ for the model in the previous exercise and interpret it in the context of the data.

b.  Calculate $RMSE$ for the model from the previous exercise and interpret it in the context of the data.

c.  Comment on the model fit based on $R^2$ and $RMSE$.

## Exercise 10

a\. Interpret the slope in the context of the data.

b\. A lemur named Strawberry (`dlc_id = 6582`) weighed 475.0 grams when she was 3.19 months old. What was predicted weight for this lemur based on the model? What is the residual?

# Submission

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember -- you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.

If you write your responses to conceptual exercises by hand, you will need to combine your written work to the completed PDF for the applied exercises before submitting on Gradescope.

Instructions to combine PDFs:

-   Preview (Mac): [support.apple.com/guide/preview/combine-pdfs-prvw43696/mac](https://support.apple.com/guide/preview/combine-pdfs-prvw43696/mac)

-   Adobe (Mac or PC): [helpx.adobe.com/acrobat/using/merging-files-single-pdf.html](https://helpx.adobe.com/acrobat/using/merging-files-single-pdf.html)

    -   Get free access to Adobe Acrobat as a Duke student: [oit.duke.edu/help/articles/kb0030141/](https://oit.duke.edu/help/articles/kb0030141/)
:::

To submit your assignment:

-   Access Gradescope through the menu on the [STA 221 Canvas site.](https://canvas.duke.edu/courses/51767)

-   Click on the assignment, and you’ll be prompted to submit it.

-   Mark the pages associated with each exercise. All of the pages of your lab should be associated with at least one question (i.e., should be “checked”).

-   Select the first page of your .PDF submission to be associated with the *“Workflow & formatting”* section.

# Grading (50 points)

| Component             | Points |
|-----------------------|--------|
| Ex 1                  |        |
| Ex 2                  |        |
| Ex 3                  |        |
| Ex 4                  |        |
| Ex 5                  |        |
| Ex 6                  |        |
| Ex 7                  |        |
| Ex 8                  |        |
| Ex 9                  |        |
| Ex 10                 |        |
| Workflow & formatting |        |

The "Workflow & formatting" grade is to assess the reproducible workflow and document format for the applied exercises. This includes having at least 3 informative commit messages, a neatly organized document with readable code and your name and the date updated in the YAML.
