---
title: "AE 03: Inference"
subtitle: "NCAA Football Expenditures"
date: "February 5, 2026"
editor: visual
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta221-sp26) and locate your `ae-01` repo to get started.
:::

# Set up

```{r}
#| label: load-packages-data
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)

football <- read_csv("data/ncaa-football-exp.csv")
```

## Data: NCAA football expenditures

Today's data come from [Equity in Athletics Data Analysis](https://ope.ed.gov/athletics/#/datafile/list) and includes information about sports expenditures and revenues for colleges and universities in the United States. This data set was featured in a [March 2022 Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-29/readme.md).

We will focus on the 2019 - 2020 season expenditures on football for institutions in the NCAA - Division 1 FBS. The variables are :

-   `total_exp_m`: Total expenditures on football in the 2019 - 2020 academic year (in millions USD)

-   `enrollment_th`: Total student enrollment in the 2019 - 2020 academic year (in thousands)

-   `type`: institution type (Public or Private)

# Regression model

```{r}
exp_fit <- lm(total_exp_m ~ enrollment_th + type, data = football)

tidy(exp_fit)|> 
  kable(digits = 3)
```

## Part 1: Distribution of $\hat{\boldsymbol{\beta}}$ 

Our ultimate goal is to use statistical inference to draw conclusions about the relationship between student enrollment and spending on football programs, after accounting for the institution type.

## Exercise 1

We will use the vector of responses $\mathbf{y}$ and the design matrix $\mathbf{X}$ to compute the values needed for inference.

Get $\mathbf{y}$ and $\mathbf{X}$ from the football data frame. What are their dimensions?

## Exercise 2

Use $\mathbf{y}$ and $\mathbf{X}$ compute the estimated coefficients $\hat{\boldsymbol{\beta}}$.

## Exercise 3

-   Compute the residuals using the vector / matrix notation. You do not need to print the residuals.

-   Use the residuals to compute $\hat{\sigma}^2_{\epsilon}$.

## Exercise 4

-   Compute $Var(\hat{\boldsymbol{\beta}})$ . What are its dimensions.

-   Explain how the standard error for $\beta_j$, the coefficient of `enrollment_th` is computed from $Var(\hat{\boldsymbol{\beta}})$.

# Hypothesis test

We want to conduct a hypothesis test to determine if there is a linear relationship between enrollment and football expenditures after accounting for institution type.

We'll start by getting estimates for statistics we'll need for inference.

## Exercise 3

Now we're ready to conduct the hypothesis test between enrollment and football expenditures. State the null and alternative hypotheses in words and using mathematical notation.

## Exercise 4

Calculate $SE(\hat{\beta}_j)$, then use this value to calculate the test statistic for the hypothesis test.

## Exercise 5

Now we need to calculate p-value to help make our final conclusion.

-   State the distribution used to calculate the p-value.

-   Fill in the code below to calculate the p-value. Remove `#| eval: false` once you've filled in the code.

```{r}
#| label: ex-5
#| eval: false

2 * pt([test-statistic], [df], lower.tail = FALSE)

```

## Exercise 6

State your conclusion in the context of the data. Use a threshold of $\alpha = 0.05$.

::: callout-important
# Submission

To submit the AE:

Render the document to produce the PDF with all of your work from today's class.

Push all your work to your AE repo on GitHub. You're done! 🎉
:::
