---
title: "Lab 05: Expanding Multiple Linear Regression"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
format: html
bibliography: references.bib
---

::: callout-important
# Due date

This lab is due on **Tuesday, March 3 at 11:59pm.** To be considered on time, the following must be done by the due date:

-   Final `.qmd` and `.pdf` files pushed to your team's GitHub repo
-   Final `.pdf` file submitted on Gradescope
:::

# Introduction

In this lab, you will apply what you've learned about linear regression to analyze data on expenditures and revenues for collegiate athletic programs. The analysis will focus on working with and interpreting categorical predictors and interaction terms. You will also explore model diagnostics and assess the appropriateness of the model for the data.

# Learning goals

By the end of the lab you will be able to…

-   use multiple linear regression to analyze complex real-world data.

-   use variance-stabilizing transformations.

-   identify and address multicollinearity.

-   evaluate whether a model is appropriate for the data.

# Getting started

-   Go to the [sta221-sp26](https://www.github.com/sta221-sp26) organization on GitHub. Click on the repo with the prefix **lab-05**. It contains the starter documents you need to complete the lab.

-   Clone the repo and start a new project in RStudio. See the [Lab 00 instructions](../labs/lab-00.html) for details on cloning a repo and starting a new project in R.

-   Each person on the team should clone the repository and open a new project in RStudio. Throughout the lab, each person should get a chance to make commits and push to the repo

# Packages

You will use the following packages in today’s lab. Add other packages as needed.

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)
```

# Data

Today's data contains expenditures, revenues, and student body composition, other information about NCAA Division 1 athletic programs at colleges and universities in the United States. The data were featured as part of the [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-29/readme.md) data visualization challenge. It was originally obtained from [Equity in Athletics Data Analysis](https://ope.ed.gov/athletics/#/datafile/list), a reporting tool by the Office of Postsecondary Education of the U.S. Department of Education.

We will investigate whether institution features affect the revenues of athletics programs in different sports. Here is a table describing the variables in the data.

This analysis will focus on the following variables:

-   `institution_name`: Institution name

-   `sector_name`: Institution type

-   `ef_total_count`: Total student enrollment for binary male/female gender

-   `sum_partic_men`: Total participation in athletics for men

-   `sum_partic_women`: Total participation in athletics for women

-   `sports`: Sports name

-   `total_exp_menwomen`: Total expenditure for both (in US dollars)

-   `total_rev_menwomen`: Total revenue for all sports (in US dollars)

[Click here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-29/readme.md) for the full data dictionary.

```{r}
#| label: load-data
#| echo: true
#| message: false
#| warning: false
sports <- read_csv("data/ncaa-revenues.csv")
```

# Exercises

**Goal**: The goal of the analysis is to use various features of NCAA Division 1 colleges and universities to explain variability in the total revenue from their athletic programs.

## Exercise 1

We'll begin by rescaling the response and some of the predictor variables.

-   Rescale the response variable `total_rev_menwomen` and the primary predictor variable `total_exp_menwomen`, so that they are in terms of \$100K (100 thousand dollars). Name the new variables `rev100k` and `exp100k`, respectively.

-   Rescale the predictor variable `ef_total_count` , so it is in terms of thousands of students. Name the new variable `students_1k`.

-   Briefly explain why we might we rescale these variables instead of using them in the original units.

## Exercise 2

Before modeling, let's do some exploratory data analysis.

-   Make a visualization of the relationship between revenue and expenditures. Use the plot to describe the relationship between the two variables.
-   Higher values of both expenditures and revenues are associated with greater variability. Transform both variables using a `log` transformation, to deal with the potential violation of an assumption of linear regression. Name the variables `log_exp` and `log_rev`, respectively.
-   Larger values of expenditures and revenues seem to follow a slightly different trend and are associated with two sports - football and basketball. Create an indicator variable that takes value `1` if the sport is basketball or football and `0` otherwise. Name the variable `bball_football`.

## Exercise 3

-   Visualize the relationship between `log_rev` and `log_exp`.

-   From the visualization, you may notice a lot of observations form a straight diagonal line, indicating a perfect one-to-one relationship between expenses and revenues. Provide a possible interpretation of this phenomenon.

-   Do you think it is reasonable to include observations displaying this exact relationship? Briefly explain.

-   Create a new data frame filtering out the observations for which expenditures and revenues are exactly equal. Call the new data frame `sports_nolinear`.

**You will use `sports_nolinear` for the remainder of the assignment.**

```{r}
#| include: false
sports <- sports |> 
  mutate(bball_football = 
           as.factor(if_else(sports == "Basketball" | sports== "Football", 1, 0)),
        exp_100k = total_exp_menwomen / 100000,
        rev_100k = total_rev_menwomen / 100000,
         log_exp = log(exp_100k),
         log_rev = log(rev_100k))

sports_nolinear <- sports |>
  filter(log_exp != log_rev)
```

## Exercise 4

Use `sports_nolinear` to fit a regression model with the log-transformed revenue as the response variable and the following predictors: log-transformed expenditures, student enrollment, institution type, sports type, participation in athletics for men, participation in athletics for women, the basketball/football indicator you created in a previous exercise, and the interaction between the log-transformed expenditures and the basketball/football indicator.

Neatly display the model using 3 digits.

## Exercise 5

Consider the regression model from the previous exercise.

-   What type of institution and sport correspond to the intercept?

-   You'll notice that one coefficient has a missing value. Why is the coefficient missing? What is the technical name of this phenomenon?

## Exercise 6

For the sake of interpretability, it is useful to have a regression model in which no coefficients are missing, and the coefficients for each sport indicator represent the baseline level for such sport. To address this issue, use the code below to fit another regression model that uses the same predictors as before, making sure to drop the unnecessary variables and the intercept (by the the `-1` in the formula) to achieve this.

```{r}
#| echo: true
sports_fit_2 <- lm(log_rev ~ -1 + sports + students_1k + sector_name +
             sum_partic_men + sum_partic_women + log_exp + 
             log_exp*bball_football - bball_football,
          data = sports_nolinear)
```

-   Why is there only one coefficient for institution type, even after the intercept was removed?

-   Which type of institution was chosen to be the baseline in this model?

## Exercise 7

Now that we have an interpretable model, let us assess the model fit and perform diagnostics to verify whether our linear regression assumptions are reasonable for this data.

As a first step, provide some overall measures of model fit and comment on whether it seems to have an acceptable predictive power on the response of interest.

## Exercise 8

Next, let's take a look at the residuals for the model.

Because there are multiple sports at every institution, we may be concerned the residuals within an institution are correlated with each other (thus violating the independence assumption).

Due to the large number of institutions, we will look at randomly selected subset of 20 institutions to evaluate this.

-   Take a random sample of 20 institutions. Use `set.seed(221)` to make your results reproducible.

-   Plot the residuals versus fitted values, faceted by institution.

-   Based on the faceted plot, do the errors appear to be correlated within institutions? Briefly explain your response.

::: callout-tip
[Click here](https://r-graph-gallery.com/223-faceting-with-ggplot2.html) for more detail, code, and examples of faceting in ggplot2.

You may need to change the size of the figure so that the faceted lot is fully visible. You can do so using the options `#| fig-width` and `#| fig-height` in the code chunk.
:::

## Exercise 9

After an examination of all the residuals, we notice a few things:

1.  There seems to be two groups of observations.
2.  Institutions with larger fitted values correspond to lower variance in the residuals compared to the others.
3.  There are institutions with lower fitted values that have large negative residuals, potentially indicating outliers and/or influential points.

We are concerned that these observations may indicate some model misspecification (i.e., the model does not accurately reflect the trends in the data). Therefore, we take a look at the residuals a different way. We plot the standardized residual versus the fitted values, color the points based Cook's distance, and use shape to indicate whether the sport is basketball or football.

```{r}
#| echo: false

library(tidymodels)
sports_aug <- augment(sports_fit_2)
ggplot(sports_aug) +
  geom_point(mapping = aes(x = .fitted,
                           y = .std.resid,
                           color = .cooksd,
                           shape = as.factor(bball_football)))
```

-   Describe what you observe from the plot and how your observations compare to the list above.

-   Do you think this model is an appropriate fit for the data or is the model misspecified? Briefly explain.

## Exercise 10

-   Based on this model (regardless of your answer to the previous exercise), which variables seem to be useful in explaining the variability in the revenue from collegiate sports?

-   Interpret the coefficient for one useful quantitative predictor in terms of the log-transformed revenue.

-   Interpret the coefficient for one useful categorical predictor in terms of the log-transformed revenue.

# Submission

You will submit the PDF documents for labs, homework, and exams in to Gradescope as part of your final submission.

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember – you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.
:::

To submit your assignment:

-   Access Gradescope through the menu on the [STA 221 Canvas site.](https://canvas.duke.edu/courses/73984)

-   Click on the assignment, and you’ll be prompted to submit it.

-   Select the name of **every** team member.

-   Mark the pages for the *"Exercises"* question.

-   Select the first page of your .PDF submission to be associated with the *"Team agreement"* and *“Workflow & formatting”* questions.

# Grading (10 pts)

This lab will be graded based on completion and workflow & formatting. The point breakdown is as follows:

| Component             | Points |
|-----------------------|--------|
| Complete exercises    | 8      |
| Workflow & formatting | 2      |

: **Completion:** Points will be awarded for completion based on the following:

-   8 pts: Complete all exercises

-   7 pts: Complete 8 - 9 exercises

-   6 pts: Complete 6 - 7 exercises

-   5 pts: Complete 4 - 5 exercises

-   4 pts: Complete 2 - 3 exercises

-   0 pts: Complete \< 2 exercises

**Workflow & formatting**

The “Workflow & formatting” grade is to assess the reproducible workflow and collaboration. This includes having at least one meaningful commit from each team member and updating the team name and date in the YAML.

-   2 pts: Meet all criteria

-   1 pt: Meet some criteria

-   0 pt: Meet no criteria
