---
title: "Lab 07: Logistic Regression"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
format: html
bibliography: references.bib
---

::: callout-important
# Due date

This lab is due on **Tuesday, April 9 at 11:59pm.** To be considered on time, the following must be done by the due date:

-   Final `.qmd` and `.pdf` files pushed to your team's GitHub repo
-   Final `.pdf` file submitted on Gradescope
:::

# Logistic Regression

In this lab, you will use logistic regression analysis to classify rice varieties based on features derived from image data. You will fit and interpret the logistic regression models using training data, then use testing data to assess how well the model classifies observations into the two varieties.

# Learning goals

By the end of the lab you will be able to…

-   conduct exploratory data analysis for data with a binary response variable.

-   fit and interpret logistic regression model.

-   use relevant metrics and subject-matter context to identify a threshold for classification.

-   use training and testing data to fit and assess model performance.

# Getting started

-   Go to the [sta221-sp26](https://www.github.com/sta221-sp26) organization on GitHub. Click on the repo with the prefix **lab-07**. It contains the starter documents you need to complete the lab.

-   Clone the repo and start a new project in RStudio. See the [Lab 00 instructions](../labs/lab-00.html) for details on cloning a repo and starting a new project in R.

-   Each person on the team should clone the repository and open a new project in RStudio. Throughout the lab, each person should get a chance to make commits and push to the repo.

# Packages

You will use the following packages in today’s lab. Add other packages as needed.

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)
library(pROC)
```

# Data

The data in this lab contains measures describing the shape and structure of two rice varieties - *Cammeo* and *Osmancik*. To curate the data set, researchers used images from over 3,000 grains a rice in these two varieties. They then used automated methods to process the image data and extract 7 morphological features (features related to the structure) from each image. The data were originally presented and analyzed in @cinar2019classification and was obtained from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/545/rice+cammeo+and+osmancik).

This analysis will focus on the following variables:

-   `Class`: Cammeo or Osmancik

-   `Area`: Size of the rice grain measured in pixels

-   `Eccentricity`: A measure of how round the ellipse is, i.e., how close the shape of the grain is to a circle.

[Click here](https://archive.ics.uci.edu/dataset/545/rice+cammeo+and+osmancik) for the full data dictionary.

```{r}
#| lable: read-data
rice <- read_csv("data/rice.csv")
```

# Exercises

**Goal**: The goal of the analysis is to use Area and Eccentricity to identify grains from the Cammeo variety versus those from the Osmancik variety.

## Exercise 1

We'll begin by exploring the data. Create a scatterplot of `Area` versus `Eccentricity` with the color and shape of the points by `Class`.

## Exercise 2

Based on the plot from the previous exercise, do you think the two types of rice can be distinguished based on `Area` and `Eccentricity`. Briefly explain.

## Exercise 3

With this type of classification problem, it is common to test the performance of a logistic regression model by splitting the data into a *training* and a *test* set. This means that we will choose a random subset of the data to estimate the regression coefficient (training), and we will then test the predictions on the remaining observations (test).

-   Randomly select 75% of the rows of the data to create the training set `rice_train` . Use `set.seed(221)`.

-   Put the remaining 25% of the observations in the testing set called `rice_test`.

## Exercise 4

In a logistic regression model, the *log-odds* of the response being "1" (or a "success") is given by $$\log\Big(\frac{\pi_i}{1-\pi_i}\Big) = \mathbf{x}_i^\mathsf{T} \boldsymbol{\beta}$$

In this analysis a "success" is the Class "Osmancik".

-   What does each $\pi_i$ represent in the context of this analysis?

-   How is the probability of the response variable being "1" calculated from the log-odds? Show the mathematical steps to go from the log-odds to the probability in your response.

## Exercise 5

Use `rice_train` to fit the logistic regression model for the response variable `Class` on the predictors `Area` and `Eccentricity` .

-   Neatly display the output using 3 digits.
-   Does the intercept have any reasonable interpretation? If so, interpret the intercept. Otherwise explain why not.

## Exercise 6

-   Interpret the coefficients on `Area` and `Eccentricity` in the context of the data in terms of the odds and comment on whether they are useful predictors of `Class` in the model.

## Exercise 7

-   How would you expect the log-odds of the rice grain being of the Osmancik variety to change if the measures eccentricity changes from 0.85 to 0.9?

-   How would you expect the odds of the rice grain being of the Osmancik variety to change if the measures eccentricity changes from 0.85 to 0.9?

## Exercise 8

Now let's test our model on the test set. Use the `predict()` function to obtain the predicted probabilities of the rice being Osmancik for the observations in `rice_test.`

## Exercise 9

With these estimated probabilities, we can now try to classify the rice in the test set. Choose a threshold for assigning a class to each observation based on the estimated probability. Briefly explain your reasoning for selecting the threshold, including any analysis used to make your decision.

## Exercise 10

Compare the estimated class assignments you constructed with the actual classes. Comment on the result and thus the model performance.

# Submission

You will submit the PDF documents for labs, homework, and exams in to Gradescope as part of your final submission.

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember – you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.
:::

To submit your assignment:

-   Access Gradescope through the menu on the [STA 221 Canvas site.](https://canvas.duke.edu/courses/73984)

-   Click on the assignment, and you’ll be prompted to submit it.

-   Select the name of **every** team member.

-   Mark the pages for the *"Exercises"* question.

-   Select the first page of your .PDF submission to be associated with the *"Team agreement"* and *“Workflow & formatting”* questions.

# Grading (10 pts)

This lab will be graded based on completion and workflow & formatting. The point breakdown is as follows:

| Component             | Points |
|-----------------------|--------|
| Complete exercises    | 8      |
| Workflow & formatting | 2      |

: **Completion:** Points will be awarded for completion based on the following:

-   8 pts: Complete all exercises

-   7 pts: Complete 8 - 9 exercises

-   6 pts: Complete 6 - 7 exercises

-   5 pts: Complete 4 - 5 exercises

-   4 pts: Complete 2 - 3 exercises

-   0 pts: Complete \< 2 exercises

**Workflow & formatting**

The “Workflow & formatting” grade is to assess the reproducible workflow and collaboration. This includes having at least one meaningful commit from each team member and updating the team name and date in the YAML.

-   2 pts: Meet all criteria

-   1 pt: Meet some criteria

-   0 pt: Meet no criteria
